I was able to complete this degree and become the researcher I am today thanks to the unwavering support of many amazing people. First off, I would like to thank my advisors, Tasha Beretvas and James Pustejovsky. I am extremely grateful for the chance to learn and work with you both. Through every challenge, you remained incredibly supportive. I could not have hoped for more knowledgeable, conscientious, and kind mentors. 

Tasha, thank you for introducing me to meta-analysis and encouraging me to apply to the Quantitative Methods program. I hope to be as thoughtful and engaging a teacher and researcher as you are. Your humor has helped me get through hard times, and I will be forever grateful to have met you and spent the past years getting to know you. 

James, thank you for taking me on as a student and welcoming me into the Pusto Lab. Beyond inspiring me to work diligently and to uphold high standards, I really value your supportive and open-minded attitude toward collaboration and research. You taught me to be more confident and to face new challenges with excitement and curiosity instead of apprehension. 

I want to thank my committee, Tiffany Whittaker, Terri Pigott, and Xiao Liu, for their support and for giving me valuable feedback throughout this process. Terri, thank you for discussing various dissertation topics with me and pointing me toward this one in particular. I have really enjoyed working on it. 

Thank you, Mikkel, for collaborating with me during your dissertation. Extending your work has been really fun for me, and I have appreciated the thoughtfulness and ingenuity you, James, and Terri brought to \textcite{vembye2023}.

To my undergraduate advisors, Stephen Sonnenberg and Thomas Palaima, thank you for nurturing my early interest in academia. Dr. Sonnenberg, I will always fondly remember the times you took me out for coffee and read back my thesis to me. It was extremely kind, and any writing ability I have is truly attributable to your mentorship. 

I have been extremely fortunate to have made some of the best friends of my life while working on my PhD. Melissa and Megha, thank you for your fantastic feedback on this document and my defense presentation. You both have really been there for me during one of the most trying times of my life. Megha, I admire the dedication and enthusiasm you bring to your passions, and I am deeply grateful for the emotional support you have given me over the years. Melissa, I could not ask for a better dissertation buddy. Even if our meetings weren't always focused, our conversations always renewed my resolve. Saleem, thank you for giving me tips on running my simulation on TACC. I really had fun that night at your and Megha's place. 

Molly and Gleb, thank you for always being there for me across all areas of life. I feel truly fortunate to have formed such strong connections with you both. Molly, I have had so much fun going on adventures with you. You are the strongest person I know, and have been like a sister to me. Gleb, I have always felt like an odd duck, and maybe you're right -- we just happen to share the same mental pathology. Either way, I'm grateful to have met someone so similar to me. Thank you for making me laugh and helping me become a kinder, more empathetic person. Thank you, Pierce and Suhwa, for working alongside me for so many hours at cafes around town and for being the best friends I could ask for. Thank you, Brandie, Sook Hyun, Man, Chris, Danny, Daniel, Young Ri, and Jihyun, for your advice and engaging conversations.

Thank you, Io, Hema, and Cary Grant, for giving me comfort and compassion in times of isolation. Your love and presence in my life bring me a happiness that is purer than anything else that I experience. 

Thank you, mummy-ji and papa-ji, for helping me heal and take care of Nirvaan in the months following his birth. Your help was invaluable to me, and I would not have been able to finish my degree or gain my strength back so quickly without the help you gave me. Mummy-ji, your daily inquiries about my work progress (and your inquiries about whether I found the inquiries about my work progress helpful or annoying) were super sweet and a happy memory that I will hold on to. Papa-ji, your good wishes after my defense filled me with pride. 

Mom, seeing you progress through your PhD while I was in high school and undergrad, inspired me to push myself. Your resilience continues to be my model. Thank you for always supporting and guiding me. Josiah, I am lucky to have such a kind, talented, and intelligent brother. I look forward to our conversations, and I often wish for an afternoon to play Katamari with you as we did as kids. Sophie, the day you were born was a truly significant one for me, because it brought you into my life. No matter what, I know that we will always be there for each other.

Dad, losing you broke my world. Not only did I lose my father, but I lost my best friend. I do not believe I will ever meet someone as creative or genuine as you. I will try to fill the hole you left with all that you have taught and shared with me. If I can even attempt to live my life with as much passion and curiosity as you did yours, I will be satisfied. 

Abhi, thank you for always being here for me over the past 11 years. Your intelligence and intuition astound me daily. The love, support, and friendship you have given me have had a huge impact on my life, and without you, I could have done none of this. In time, I hope to give back to you all that you have given me. I love you more than anything, and I cherish every moment I get to spend with you. 

Finally, Nirvaan, thank you for hanging in there this past year. When I found out I was pregnant with you, I was overjoyed. Your papa and I have joked that you have been uncharacteristically agreeable for an infant (even coming a month early to spare your poor mama more pain). Your presence in my life gave me the strength and motivation to carry this through. I love you, and I will always be there for you. 
